---
title: "Trap Reanalysis using DESeq2"
output:
  html_document:
    keep_md: yes
    theme: united
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  fig.path = "figures/DESeq2-"
)
```

Here we reproduce the re-analysis of the data from [Marrocco J. et al. 2019](https://doi.org/10.3389/fnbeh.2019.00157) shown in the [readme](README.md), but using, like the authors, DESeq2 for the statistical analysis.

## Assessment of Marrocco et al. 2019 using DESeq2

Marrocco et al. 2019 were reporting changes in the acute stress response (forced-swim test, FST) depending on the prior exposure to early life stress (ELS). We reproduce this analysis using appropriate statistical models to handle interactions between groups:

```{r results='hide', message=FALSE, warning=FALSE}
library(SEtools)
library(SummarizedExperiment)
library(plgINS)
library(DESeq2)
library(ggplot2)
source("misc.R")
se <- readRDS("data/AllData.kallisto.SE.rds")
```

```{r}
se <- se[,se$Set=="GSE131972"]
dds1 <- DESeqDataSetFromMatrix(round(assay(se)), colData=colData(se),
        design=~FST+ELS+FST:ELS )
dds1 <- DESeq(dds1)
# so that we plot the DESeq-normalized counts:
dat1 <- se[row.names(dds1),]
assays(dat1)$lognorm <- log1p(counts(dds1, normalized=TRUE))
rowData(dat1)$logCount <- rowMeans(assays(dat1)$lognorm)
resultsNames(dds1)
```

### Effects of FST

```{r, fig.width=6, fig.height=3}
res <- results(dds1, name = "FST_FST_vs_None")
res <- res[order(res$padj),]
dat1 <- dat1[,order(dat1$ELS, dat1$FST)]
sehm(dat1, row.names(res)[which(res$padj<0.05)], assayName = "lognorm", do.scale = TRUE, anno_row="logCount", anno_columns=c("ELS","FST"), main="Differentially expressed upon FST")
```

We find a few genes which are significantly differentially expressed upon FST.

```{r, fig.width=6, fig.height=3}
res <- results(dds1, name = "ELS_ELS_vs_None")
res <- res[order(res$padj),]
dat1 <- dat1[,order(dat1$ELS, dat1$FST)]
sehm(dat1, row.names(res)[which(res$padj<0.05)], assayName = "lognorm", do.scale = TRUE, anno_row="logCount", anno_columns=c("ELS","FST"), main="Differentially expressed upon ELS")
```

We find a few genes which are differentially expressed in the ELS group with respect to the non-exposed animals.


```{r}
res <- results(dds1, name = "FSTFST.ELSELS")
res <- res[order(res$padj),]
dat1 <- dat1[,order(dat1$ELS, dat1$FST)]
res[which(res$padj<0.05),]
```

We find a single gene that is differentially expressed after FST in ELS vs non-ELS mice - considerably less than the hundreds reported by the authors.

