---
title: "Trap Reanalysis - staged FDR"
output:
  html_document:
    keep_md: yes
    theme: united
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  fig.path = "figures/stageR-"
)
```


Here we combine the data from both accessions and test for interactions, using the more lenient staged false-discovery rate approach  developed by <a href="https://doi.org/10.1186/s13059-017-1277-0">Van den Berge et al. (2017)</a>.

```{r results='hide', message=FALSE, warning=FALSE}
library(edgeR)
library(SummarizedExperiment)
library(SEtools)
library(stageR)
library(ggplot2)
source("misc.R")

se <- readRDS("data/AllData.kallisto.SE.rds")
se <- dosvacor(se, form = ~FST + Sex + Genotype + Set + ELS, form0 = ~Set)
```

### Is the response to FST significantly different between sexes?
```{r}
design <- model.matrix(~SV1+SV2+FST+Sex+Genotype+Set+ELS+FST:Sex, data=as.data.frame(colData(se)))

y <- DGEList(counts=assays(se)$counts)
y <- calcNormFactors(y)
y <- estimateDisp(y,design)
y <- y[filterByExpr(y, design),]

fit <- glmQLFit(y,design)

res1 <- glmQLFTest(fit, grep("FST|Sex", colnames(design), value=TRUE))$table
pScreen <- res1$PValue
names(pScreen) <- row.names(res1)

pConf <- as.matrix(glmQLFTest(fit, "FSTFST:Sexfemale")$table[,"PValue",drop=FALSE])
colnames(pConf) <- "FST:Sex"

stageRObj <- stageR(pScreen=pScreen, pConfirmation=pConf, pScreenAdjusted=FALSE)
stageRObj <- stageWiseAdjustment(object=stageRObj, method="none", alpha=0.05)
res <- getAdjustedPValues(stageRObj, onlySignificantGenes=FALSE, order=FALSE)
head(res <- res[order(res[,2]),])
degs <- row.names(res)[which(res[,2]<0.05)]
d1 <- data.frame(feature=degs, q=paste0("q~",format(res[degs,2], digits=2)))
```

Two genes pass FDR:

```{r, fig.width=5, fig.height=3.5}
ggplot(meltSE(se, d1$feature), aes(Sex, corrected)) + geom_boxplot(aes(fill=FST), outlier.shape=NA) + 
  ylab("Corrected logCPM") + facet_wrap(~feature, scale="free_y") + 
  geom_point(position=position_dodge(width=0.75), aes(group=FST), alpha=0.6) +
  geom_text(data=d1, aes(x=Inf, y=Inf, label=q), hjust=1.1, vjust=1.2)
```



### Is the response to FST significantly altered by BDNF Val66Met?

```{r}
design <- model.matrix(~SV1+SV2+FST+Sex+Genotype+Set+ELS+Genotype:FST, data=as.data.frame(colData(se)))

y <- DGEList(counts=assays(se)$counts)
y <- calcNormFactors(y)
y <- estimateDisp(y,design)
y <- y[filterByExpr(y, design),]

fit <- glmQLFit(y,design)

res1 <- glmQLFTest(fit, grep("Genotype|FST", colnames(design), value=TRUE))$table
pScreen <- res1$PValue
names(pScreen) <- row.names(res1)

pConf <- as.matrix(glmQLFTest(fit, "FSTFST:GenotypeBDNFMET")$table[,"PValue",drop=FALSE])
colnames(pConf) <- "Genotype:FST"

stageRObj <- stageR(pScreen=pScreen, pConfirmation=pConf, pScreenAdjusted=FALSE)
stageRObj <- stageWiseAdjustment(object=stageRObj, method="none", alpha=0.05)
res <- getAdjustedPValues(stageRObj, onlySignificantGenes=FALSE, order=FALSE)
head(res <- res[order(res[,2]),])
degs <- row.names(res)[which(res[,2]<0.5)]
d2 <- data.frame(feature=degs, q=paste0("q~",format(res[degs,2], digits=2)))
```

A single gene passes FDR:

```{r, fig.width=7, fig.height=3.5}
ggplot(meltSE(se, d2$feature), aes(Genotype, corrected)) + geom_boxplot(aes(fill=FST), outlier.shape=NA) + 
  ylab("Corrected logCPM") + facet_wrap(~feature, scale="free_y") + 
  geom_point(position=position_dodge(width=0.75), aes(group=FST), alpha=0.6) +
  geom_text(data=d2, aes(x=Inf, y=Inf, label=q), hjust=1.1, vjust=1.2)
```

### Is the response to FST significantly altered by early life stress?
```{r}
design <- model.matrix(~SV1+SV2+FST+Sex+Genotype+Set+ELS+ELS:FST, data=as.data.frame(colData(se)))

y <- DGEList(counts=assays(se)$counts)
y <- calcNormFactors(y)
y <- estimateDisp(y,design)
y <- y[filterByExpr(y, design),]

fit <- glmQLFit(y,design)

res1 <- glmQLFTest(fit, grep("ELS|FST", colnames(design), value=TRUE))$table
pScreen <- res1$PValue
names(pScreen) <- row.names(res1)

pConf <- as.matrix(glmQLFTest(fit, "FSTFST:ELSELS")$table[,"PValue",drop=FALSE])
colnames(pConf) <- "ELS:FST"

stageRObj <- stageR(pScreen=pScreen, pConfirmation=pConf, pScreenAdjusted=FALSE)
stageRObj <- stageWiseAdjustment(object=stageRObj, method="none", alpha=0.05)
res <- getAdjustedPValues(stageRObj, onlySignificantGenes=FALSE, order=FALSE)
head(res <- res[order(res[,2],res[,1]),])
```

No gene passes multiple testing correction.
