---
title: "Trap Reanalysis"
output:
  html_document:
    keep_md: yes
    theme: united
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  fig.path = "README_figs/README-"
)
```

required r libraries
```{r results='hide', message=FALSE, warning=FALSE}
library(edgeR)
library(SEtools)
library(SummarizedExperiment)
library(plgINS)
library(DESeq2)
library(sva)
```


# About this re-analysis

Here, we outline the re-analysis of two GEO datasets, GSE100579 and GSE131972, both of which are investigating the effect of acute stress on the translatome of CA3 pyramidal neurons in the mouse hippocampus.

This repositories were used for 3 seperate publications:

Marrocco J. et al. 2017; 
A sexually dimorphic pre-stressed translational signature in CA3 pyramidal neurons of BDNF Val66Met mice; Nature Communications volume 8, Article number: 808 (2017)

Gray J. D. et al. 2018; 
Translational profiling of stress-induced neuroplasticity in the CA3 pyramidal neurons of BDNF Val66Met mice; Molecular Psychiatry volume 23, pages 904–913

Marrocco J. et al. 2019; 
Early Life Stress Restricts Translational Reactivity in CA3 Neurons Associated With Altered Stress Responses in Adulthood; Front. Behav. Neurosci. https://doi.org/10.3389/fnbeh.2019.00157 


the reanalysis be performed on the quantification of two different alginment methods a genome alignment using xxx and salmon or a pseudoalignment to the transcriptome using kallisto

first, we are loading both the kallisto and the genome alignment data

```{r}
kallistodata <- readRDS("data/AllData.kallisto.SE.rds")
salmondata <- readRDS("data/GSE131972.salmonv2.SE.rds")
```

# Re-analysis of published results
## Assessment of Marrocco et al. 2017

Concerning the 2017 publication the authors claim that numerous genes are differentially regulated between males and females after acute stress. Unfortunately, they do not include a list with all genes in the publication, but Table 1 contains a subsets of genes that are allegedly differentially regulated between males and females upon acute stress

Let's inspect these genes across the runs used for the original publication in WildType animals

```{r}
se <- kallistodata
se <- subset(se,select  = se$Set == "GSE100579" & se$Genotype == "WildType")
genes <- read.table("metadata/Marrocco2017SexStressGenes.csv", sep = ";", header = T)$genes
sehm(se,genes,do.scale = T,anno_columns = c("ELS","FST","Genotype","Sex","Set"), cluster_rows = T)
```

It looks as if there could be something to this, however unfortunately this finding is higly underpowered having only 6 replicates for a 4 group comparison

Let's inspect these genes across all sequencing runs and find out wheather the finding for male WildType mice in the second set GSE131972 holds up.

```{r}
se <- kallistodata
se <- subset(se,select  = se$ELS == "None" & se$Genotype == "WildType")
sehm(se,genes,do.scale = T,anno_columns = c("ELS","FST","Genotype","Sex","Set"), cluster_rows = T)
```

it becomes apparent that these genes seem to be co-expressed and highly variable across all samples. the finding that in females and males these genes react differently to stress seems inaccurate. It highlights the danger of applying single run sequencing.

If we further visualize these genes across all samples, we find
```{r}
se <- kallistodata
sehm(se,genes,do.scale = T,anno_columns = c("ELS","FST","Genotype","Sex","Set"), cluster_rows = T)
```

The selected genes display a high co-expression and could possibly be highly variable due to technical reasons. Let's try to remove technical variability from our runs using SVA and re-visualize the same genes

```{r}
#function for SVA correction
dosvacor <- function(SE, form=NULL, form0=~1, ...){
  CD <- as.data.frame(colData(SE))
  mm <- model.matrix(form, data=CD)
  mm0 <- model.matrix(form0, data=CD)

  dds <- DESeqDataSetFromMatrix(round(assay(SE)), as.data.frame(colData(SE)), form)
  dds <- estimateSizeFactors(dds)
  en <- as.matrix(assay(vst(dds, blind=FALSE)))

  sv <- sva(en, mm, mm0, n.sv=NULL, ...)
  n.sv <- sv$n.sv
  sv <- sv$sv
  
  colnames(sv) <- paste0("SV",1:ncol(sv))
  X <- cbind(mm, sv)
  mm2 <- cbind(mm[,1,drop=F],sv,mm[,-1,drop=F])
  H <- solve(t(X)%*%X)%*%t(X)
  b <- (H%*%t(en))
  cn <- setdiff(colnames(X),setdiff(colnames(mm), colnames(mm0)))  
  cn <- setdiff(cn, "(Intercept)")
  encor <- en - t(as.matrix(X[,cn]) %*% b[cn,])
  SE <- SE[row.names(encor),]
  colData(SE) <- cbind(colData(SE), sv)
  assays(SE)$corrected <- encor
  return(SE)
}

se <- dosvacor(se, form = ~ Set + Sex * FST, form0 = ~Set)
sehm(se,genes,do.scale = T,anno_columns = c("ELS","FST","Genotype","Sex","Set"), cluster_rows = T,assayName = "corrected")
```

As we can see, removing technical variability abolishes the effects for the male-female stress effects seen in the single replicate comparisons. It becomes apparent, that these genes do not have any sex specific acute stress response across all samples

## Assessment of Gray et al. 2018

A complete assessment of the findings of Gray et al. 2018 is unfortunately not possible. the repository GSE100579 is missing crucial samples for the chronic stress model for the BDNF Val66Met genotype and only includes samples of acute stress.

However, the authors claim that many genes are differentially regulated between WildType and BDNF Val66Met animals at baseline. While no complete list of genes has been included in the publication, a subset can be found in Table 1

```{r}
se <- kallistodata
genes <- read.table("metadata/Gray2018GenotypeGenes.csv", sep = ";", header = T)$genes
se <- subset(se,select  = se$Set == "GSE100579" & se$FST == "None")
sehm(se,genes,do.scale = T,anno_columns = c("ELS","FST","Genotype","Sex","Set"), cluster_rows = T)
```

Let's investigate these genes across all samples
```{r}
se <- kallistodata
sehm(se,genes,do.scale = T,anno_columns = c("ELS","FST","Genotype","Sex","Set"), cluster_rows = T)
```

We see a similar pattern as in Marrocco et al 2017. The baseline difference does not reproduce in GSE 131972 which only includes WildType mice.

Let's again try to eliminate technical variabilty and re-visualize the same genes

```{r}
se <- dosvacor(se, form = ~Set + Sex + Genotype + FST + ELS, form0 = ~Set)
sehm(se,genes,do.scale = T,anno_columns = c("ELS","FST","Genotype","Sex","Set"), cluster_rows = T,assayName = "corrected")
```

It becomes apparent, that these genes do not have any genotype specific baseline difference across all samples

## Assessment of Marrocco et al. 2019
Here we re-analyse Marrocco et al. 2019 checking for ELS dependent changes in the acute stress response. The same analysis will be performed twice, once using gene counts from a transcriptome alignment, and once using gene counts from a genome alignment. Both analyses had similar results

### Using kallisto data
```{r}
se <- kallistodata
se <- subset(se,select  = se$Set == "GSE131972")
```

In their publications the authors unfortunately do not upload a list with differentially expressed genes. However, in their discussion they mention a number of genes that they claim are differentially expressed between ELS and non-ELS mice after acute stress.
They claim that acute stress reduces the expression of Grin1, Grin2a, Gabbr2, and Gabra1 in CA3 neurons of non-ELS mice, but not ELS mice:
```{r}
sehm(se,c("Grin1","Grin2a","Gabbr2","Gabra1"),do.scale = T,anno_columns = c("ELS","FST"), cluster_rows = T)
```

it becomes apparent, that this finding does not look significant

Further, they claim that there is restricted list of genes selectively induced by AS in ELS mice (Per1, Npy, Nfkbia, Penk,Dusp1, Cst3, Trib1, Htra1, Sdc4, Plekhf1) but not non-ELS mice.
```{r}
sehm(se,c("Per1", "Npy", "Nfkbia", "Penk","Dusp1", "Cst3", "Trib1", "Htra1", "Sdc4", "Plekhf1"),do.scale = T,anno_columns = c("ELS","FST"), cluster_rows = T)
```

it becomes apparent, that this finding also does not look sigificant. while these genes might be increased in expression in ELS mice it is hard to see this being a significant finding, especially, since the same effect can be observed for one replicate in the non-ELS group.

The authors claim that there are a number of genes that appear to be induced by AS in both ELS and non ELS mice. these include (Egr1/2/4, Arc, Fos, and Fosb)
```{r}
sehm(se,c("Egr1", "Egr2", "Egr4", "Arc","Fos", "Fosb"),do.scale = T,anno_columns = c("ELS","FST"), cluster_rows = T)
```

For these genes it looks as if their finding might be significant.

Let's investigate the whole data with a statistical approach using egeR and a GLM type of analysis for swim effects, early life effects and interactions
```{r}
#experimental design, interactive model
design <- model.matrix(~se$FST * se$ELS)

y <- DGEList(counts=assays(se)$counts)
y <- calcNormFactors(y)
y <- estimateDisp(y,design)

#filter out genes that are below 10 counts in more than 75% of samples
keep <- rowSums(y$counts>10) >= 3
y <- y[keep, , keep.lib.sizes=FALSE]

Results <- list()
fit <- glmQLFit(y,design)
for(i in colnames(design)[-1]){
  Results[[i]] <- glmQLFTest(fit, i)
}
```

Les's investigate if there are any genes altered by acute stress
```{r}
topTags(Results$`se$FSTFST`)
```

Even though the data looked promising unfortunately nothing survives mutliple testing correction. This indicates that this study might be severly underpowered, so in order to resolve FST specific effects more replicats would be needed


Are there any genes altered by early life stress?
```{r}
topTags(Results$`se$ELSELS`)
```

Unfortunately, there are no genes that pass the multiple testing correction

Let's investigate if there are genes with a significant interaction
```{r}
topTags(Results$`se$FSTFST:se$ELSELS`)
```

Unfortunately, no genes have a altered acute stress response in ELS vs normal animals

### Using salmon data

To demonstrate that alignment to the genome results in similar results in this section we repeat the same analysis with the salmon data
```{r}
se <- salmondata
```

In their Publications the authors unfortunately do not upload a list with differentially expressed genes. However, in their discussion they mention a number of genes that they claim are differentially expressed between ELS and non-ELS mice after acute stress.
They claim that acute stress reduces the expression of Grin1, Grin2a, Gabbr2, and Gabra1 in CA3 neurons of non-ELS mice, but not ELS mice:
```{r}
sehm(se,c("Grin1","Grin2a","Gabbr2","Gabra1"),do.scale = T,anno_columns = c("ELS","FST"), cluster_rows = T)
```

it becomes apparent, that this finding does not look significant

Further, they claim that there is restricted list of genes selectively induced by AS in ELS mice (Per1, Npy, Nfkbia, Penk,Dusp1, Cst3, Trib1, Htra1, Sdc4, Plekhf1) but not non-ELS mice.
```{r}
sehm(se,c("Per1", "Npy", "Nfkbia", "Penk","Dusp1", "Cst3", "Trib1", "Htra1", "Sdc4", "Plekhf1"),do.scale = T,anno_columns = c("ELS","FST"), cluster_rows = T)
```

it becomes apparent, that this finding also does not look significant. while these genes might be increased in expression in ELS mice it is hard to see this being a significant finding, especially, since the same effect can be observed for one replicate in the non-ELS group.

The authors claim that there are a number of genes that appear to be induced by AS in both ELS and non ELS mice. these include (Egr1/2/4, Arc, Fos, and Fosb)
```{r}
sehm(se,c("Egr1", "Egr2", "Egr4", "Arc","Fos", "Fosb"),do.scale = T,anno_columns = c("ELS","FST"), cluster_rows = T)
```

For these genes it looks as if their finding might be significant.

Let's investigate the whole data with a statistical approach using egeR and a GLM type of analysis for swim effects, early life effects and interactions
```{r}
#experimental design, interactive model
design <- model.matrix(~se$FST * se$ELS)

y <- DGEList(counts=assays(se)$counts)
y <- calcNormFactors(y)
y <- estimateDisp(y,design)

#filter out genes that are below 10 counts in more than 75% of samples
keep <- rowSums(y$counts>10) >= 3
y <- y[keep, , keep.lib.sizes=FALSE]

Results <- list()
fit <- glmQLFit(y,design)
for(i in colnames(design)[-1]){
  Results[[i]] <- glmQLFTest(fit, i)
}
```

Les's investigate if there are any genes altered by acute stress
```{r}
topTags(Results$`se$FSTFST`)
```

Even though the data looked promising unfortunately nothing survives mutliple testing correction. This indicates that this study might be severly underpowered, so in order to resolve FST specific effects more replicats would be needed.


Are there any genes altered by early life stress?
```{r}
topTags(Results$`se$ELSELS`)
```

Unfortunately, there are no genes that pass the multiple testing correction

Let's investigate if there are genes with a significant interaction
```{r}
topTags(Results$`se$FSTFST:se$ELSELS`)
```

Unfortunately, no genes have a altered acute stress response in ELS vs normal animals

# Meta-Analysis of all data

## Additive model
Let's run an overarching analysis over all data to determine if there are any significant effects for acute stress (=FST), Genotype, GEOdataset or early life stress
```{r}
se <- kallistodata

#experimental design, full additive model
design <- model.matrix(~se$FST + se$Sex + se$Genotype + se$Set + se$ELS)

y <- DGEList(counts=assays(se)$counts)
y <- calcNormFactors(y)
y <- estimateDisp(y,design)

#filter out genes that are below 10 counts in more than 75% of samples
keep <- rowSums(y$counts>10) >= 5
y <- y[keep, , keep.lib.sizes=FALSE]

Results <- list()
fit <- glmQLFit(y,design)
for(i in colnames(design)[-1]){
  Results[[i]] <- glmQLFTest(fit, i)
}
```

Les's investigate if there are any genes altered by acute stress
```{r}
topTags(Results$`se$FSTFST`)
```
Indeed, there are multiple candidate genes that are significantly altered by acute stress across other Conditions

Are there any genes altered by sex
```{r}
topTags(Results$`se$Sexfemale`)
```
Indeed, there are multiple candidate genes that are significantly altered by sex across other FSTs


Are there any genes altered by BDNF Val66Met phenotype?
```{r}
topTags(Results$`se$GenotypeBDNFMET`)
```

Unfortunately, there are no genes that pass the multiple testing correction

Are there any genes altered early life stress?
```{r}
topTags(Results$`se$ELSELS`)
```

Unfortunately, there are no genes that pass the multiple testing correction

Let's plot the top 10 genes for both sex and acute stress
```{r}
genes <- append(rownames(topTags(Results$`se$FSTFST`)),rownames(topTags(Results$`se$Sexfemale`)))
sehm(se,genes,do.scale = T,anno_columns = c("ELS","FST","Genotype","Sex","Set"), cluster_rows = T,assayName = "logcpm")
```

## Interactive models

To futher investigate if across all samples there is any variable that interacts with acute stress we will run a series of models that incorporate an interaction term between acute stress and any of the variables

Interaction between acute stress and sex
```{r}
se <- kallistodata

#experimental design, full additive model
design <- model.matrix(~se$FST + se$Sex + se$Genotype + se$Set + se$ELS + se$FST:se$Sex)

y <- DGEList(counts=assays(se)$counts)
y <- calcNormFactors(y)
y <- estimateDisp(y,design)

#filter out genes that are below 10 counts in more than 75% of samples
keep <- rowSums(y$counts>10) >= 5
y <- y[keep, , keep.lib.sizes=FALSE]

Results <- list()
fit <- glmQLFit(y,design)
for(i in colnames(design)[-1]){
  Results[[i]] <- glmQLFTest(fit, i)
}
topTags(Results$`se$FSTFST:se$Sexfemale`)
```

Interaction between acute stress and genotype
```{r}
se <- kallistodata

#experimental design, full additive model
design <- model.matrix(~se$FST + se$Sex + se$Genotype + se$Set + se$ELS + se$FST:se$Genotype)

y <- DGEList(counts=assays(se)$counts)
y <- calcNormFactors(y)
y <- estimateDisp(y,design)

#filter out genes that are below 10 counts in more than 75% of samples
keep <- rowSums(y$counts>10) >= 5
y <- y[keep, , keep.lib.sizes=FALSE]

Results <- list()
fit <- glmQLFit(y,design)
for(i in colnames(design)[-1]){
  Results[[i]] <- glmQLFTest(fit, i)
}
topTags(Results$`se$FSTFST:se$GenotypeBDNFMET`)
```

Interaction between acute stress and early life stress
```{r}
se <- kallistodata

#experimental design, full additive model
design <- model.matrix(~se$FST + se$Sex + se$Genotype + se$Set + se$ELS + se$FST:se$ELS)

y <- DGEList(counts=assays(se)$counts)
y <- calcNormFactors(y)
y <- estimateDisp(y,design)

#filter out genes that are below 10 counts in more than 75% of samples
keep <- rowSums(y$counts>10) >= 5
y <- y[keep, , keep.lib.sizes=FALSE]

Results <- list()
fit <- glmQLFit(y,design)
for(i in colnames(design)[-1]){
  Results[[i]] <- glmQLFTest(fit, i)
}
topTags(Results$`se$FSTFST:se$ELSELS`)
```

# DESeq2 analysis


```{r}
suppressPackageStartupMessages({
  library(DESeq2)
  library(sva)
  library(SEtools)
  library(RUVSeq)
})
```

## GSE100579

```{r}
dat1 <- readRDS("data/TRAPReanalysis.SE.rds")
dat1 <- dat1[,dat1$Set=="GSE100579"]
dat1$Genotype <- factor(gsub(" ","",dat1$Geontype), c("WildType","BDNFMET"))
dat1$Geontype <- NULL
dat1$Condition <- relevel(dat1$Condition, "None")
dds1 <- DESeqDataSetFromMatrix( round(assay(dat1)), colData=colData(dat1),
        design=~Gender+Genotype+Condition+Gender:Condition )
dds1 <- DESeq(dds1)
# so that we plot the DESeq-normalized counts:
dat1 <- dat1[row.names(dds1),]
assays(dat1)$lognorm <- log1p(counts(dds1, normalized=TRUE))
rowData(dat1)$logCount <- rowMeans(assays(dat1)$lognorm)
resultsNames(dds1)
```

### Gender

```{r}
res <- results(dds1, name = "Gender_male_vs_female")
DESeq2::plotMA(res)
res <- res[order(res$padj),]
res[1:10,c(1,2,6)]
dat1 <- dat1[,order(dat1$Gender, dat1$Condition, dat1$Genotype, dat1$Estrodiol)]
sehm(dat1, row.names(res)[which(res$padj<0.05)], assayName = "lognorm", do.scale = TRUE, anno_row="logCount", anno_columns=c("Gender","Genotype","Condition"), main="Differentially-expressed across sex")
```

### FST

```{r}
res <- results(dds1, name = "Condition_FST_vs_None")
DESeq2::plotMA(res)
res <- res[order(res$padj),]
dat1 <- dat1[,order(dat1$Condition, dat1$Gender, dat1$Genotype, dat1$Estrodiol)]
sehm(dat1, row.names(res)[which(res$padj<0.05)], assayName = "lognorm", do.scale = TRUE, anno_row="logCount", anno_columns=c("Gender","Genotype","Condition"), main="Differentially-expressed upon FST")
```

Here you see exactly why I don't use DESeq: it gives significance even when there are large intra-group differences...

### Genotype

```{r}
res <- results(dds1, name = "Genotype_BDNFMET_vs_WildType")
DESeq2::plotMA(res)
res <- res[order(res$padj),]
dat1 <- dat1[,order(dat1$Genotype, dat1$Gender, dat1$Condition, dat1$Estrodiol)]
sehm(dat1, row.names(res)[which(res$padj<0.05)], assayName = "lognorm", do.scale = TRUE, anno_row="logCount", anno_columns=c("Gender","Genotype","Condition"), main="Differentially-expressed between genotypes")
```

### Gender:FST

```{r}
res <- results(dds1, name = "Gendermale.ConditionFST")
DESeq2::plotMA(res)
res <- res[order(res$padj),]
res[1:10,c(1,2,6)]
dat1 <- dat1[,order(dat1$Gender, dat1$Condition, dat1$Genotype, dat1$Estrodiol)]
sehm(dat1, row.names(res)[which(res$padj<0.05)], assayName = "lognorm", do.scale = TRUE,  anno_row="logCount", anno_columns=c("Gender","Genotype","Condition"), main="Sex-dependent effects of FST")
```

The MA-plot shows a suspicious trend. Adam17 and Prl are clearly artefacts of low counts (filtering improves both edgeR and DESeq2), especially given the trend shown by the MA-plot. RP23 is another of these weird DESeq2 calls. Kif21b and Gm7334 could be something (but see below).

# DESeq2 analysis with randomized variable reassignments

Out of interest, lets see what happens if we randomly shuffle samples and re-run the analysis


```{r}
suppressPackageStartupMessages({
  library(DESeq2)
  library(sva)
  library(SEtools)
  library(RUVSeq)
})
```

## GSE100579

```{r}
dat1 <- readRDS("data/TRAPReanalysis.SE.rds")
dat1 <- dat1[,dat1$Set=="GSE100579"]
dat1$Genotype <- factor(gsub(" ","",dat1$Geontype), c("WildType","BDNFMET"))
dat1$Geontype <- NULL
dat1$Condition <- relevel(dat1$Condition, "None")

shuffle <- sample(1:length(dat1$Condition))
dat1$Condition <- dat1$Condition[shuffle]
dat1$Genotype <- dat1$Genotype[shuffle]
dat1$Gender <- dat1$Gender[shuffle]

dds1 <- DESeqDataSetFromMatrix( round(assay(dat1)), colData=colData(dat1),
        design=~Gender+Genotype+Condition+Gender:Condition )
dds1 <- DESeq(dds1)
# so that we plot the DESeq-normalized counts:
dat1 <- dat1[row.names(dds1),]
assays(dat1)$lognorm <- log1p(counts(dds1, normalized=TRUE))
rowData(dat1)$logCount <- rowMeans(assays(dat1)$lognorm)
resultsNames(dds1)
```

### Gender

```{r}
res <- results(dds1, name = "Gender_male_vs_female")
DESeq2::plotMA(res)
res <- res[order(res$padj),]
res[1:10,c(1,2,6)]
dat1 <- dat1[,order(dat1$Gender, dat1$Condition, dat1$Genotype, dat1$Estrodiol)]
sehm(dat1, row.names(res)[which(res$padj<0.05)], assayName = "lognorm", do.scale = TRUE, anno_row="logCount", anno_columns=c("Gender","Genotype","Condition"), main="Differentially-expressed across sex")
```

### FST

```{r}
res <- results(dds1, name = "Condition_FST_vs_None")
DESeq2::plotMA(res)
res <- res[order(res$padj),]
dat1 <- dat1[,order(dat1$Condition, dat1$Gender, dat1$Genotype, dat1$Estrodiol)]
sehm(dat1, row.names(res)[which(res$padj<0.05)], assayName = "lognorm", do.scale = TRUE, anno_row="logCount", anno_columns=c("Gender","Genotype","Condition"), main="Differentially-expressed upon FST")
```

Here you see exactly why I don't use DESeq: it gives significance even when there are large intra-group differences...

### Genotype

```{r}
res <- results(dds1, name = "Genotype_BDNFMET_vs_WildType")
DESeq2::plotMA(res)
res <- res[order(res$padj),]
dat1 <- dat1[,order(dat1$Genotype, dat1$Gender, dat1$Condition, dat1$Estrodiol)]
sehm(dat1, row.names(res)[which(res$padj<0.05)], assayName = "lognorm", do.scale = TRUE, anno_row="logCount", anno_columns=c("Gender","Genotype","Condition"), main="Differentially-expressed between genotypes")
```

### Gender:FST

```{r}
res <- results(dds1, name = "Gendermale.ConditionFST")
DESeq2::plotMA(res)
res <- res[order(res$padj),]
res[1:10,c(1,2,6)]
dat1 <- dat1[,order(dat1$Gender, dat1$Condition, dat1$Genotype, dat1$Estrodiol)]
sehm(dat1, row.names(res)[which(res$padj<0.05)], assayName = "lognorm", do.scale = TRUE,  anno_row="logCount", anno_columns=c("Gender","Genotype","Condition"), main="Sex-dependent effects of FST")
```

The MA-plot shows a suspicious trend. Adam17 and Prl are clearly artefacts of low counts (filtering improves both edgeR and DESeq2), especially given the trend shown by the MA-plot. RP23 is another of these weird DESeq2 calls. Kif21b and Gm7334 could be something (but see below).


